# Numbers
I know you know what numbers are, and we've already seen that we have two types of numbers in R: integers and doubles; so, let's move to see them in action.

## Operations {-}
First of all, here is  how to do basic mathematical operation in R:
```{r}
# Plus, minus, multiply, divide
5 + 4 - 2 * 3 / 2

# Power
4 ** 3

# Logarithm
log(100) # base e
log10(100) # base 10
log2(100) # base 2
log(100, base = 3) # choose the base

# Natural exponential
exp(2)

# Square root
sqrt(9)
```

An interesting operator is the modulus (%%) which returns the remainder of a division, for example:

```{r}
7 %% 3
```

This can be useful to evaluate if a number is even or odd by calculating the remainder of the division x / 2 (so using x %% 2): if it is 0, the number x is even, otherwise it is odd.
```{r}
11 %% 2
12 %% 2
```

## Rounding {-}
Another thing that we usually want to do is to round decimal number, especially after log transformation or division. To do so, we have 3 functions:

```{r}
# Round to n decimal places
round(x = 1/3, digits = 2)

# Round to upper integer
ceiling(10.2)


# Round to lower integer
floor(14.9)
```

Look how ceiling and floor do not take into account the decimal part, even if it is greater or lower than 0.5.

## Tranform to type number {-}
Sometimes you want to transform a string that contains a number to a numeric type in R. I know we haven't covered strings yet (next chapter will be on them), but let's do a bit step forward now just to see this super useful function, that we use a lot when dealing with dataframes.
<br>
To do so, we'll use the function called `as.numeric()`
```{r}
mystring <- "15" # This is a character, can you guess why R interpret it as a character?
typeof(mystring)

mynumber <- as.numeric(mystring)
typeof(mynumber)
```
Wow, you will find this super super super useful.

## Exercises {-}
Let's now put in practice what we have seen in this chapter and in the previous (remember? the script... I want you to write a script with these exercises and save it)

::: {.exercise #mean-calc}
The results of a Real-Time PCR indicate that your triplicates for FOXP1 have these Ct: 22.4, 22.31, 22.24. Calculate the mean value and print it rounded to 2 decimal places.
:::

<details>
  <summary>Solution</summary>
```{r}
# calculator solution
mean_result_calc <- (22.4 + 22.31 + 22.24) / 3
mean_result_calc <- round(mean_result_calc, digits = 2)
print(mean_result_calc)

# BETTER solution
rep1 <- 22.4
rep2 <- 22.31
rep3 <- 22.24
n_rep <- 3

mean_res_better <- (rep1 + rep2 + rep3) / n_rep
mean_res_better_round <- round(mean_res_better, digits = 2)
print(mean_res_better_round)
```

The second solution is better because every number is stored in a variable, that you then use to calculate the mean value.
</details>


::: {.exercise #sd-calc}
Now calculate the sd of the data of exercise \@ref(exr:mean-calc) and print the value rounded to upper integer, to lower integer and to a 4-digit decimal.
:::

<details>
  <summary>Solution</summary>
```{r}
# Use only better solution, first calculate the variance
var_calc <- ((rep1 - mean_res_better)^2 + (rep2 - mean_res_better)^2 + (rep3 - mean_res_better)^2) / (n_rep - 1)

# Now let's calculate sd
sd_calc <- sqrt(var_calc)

sd_calc_ceil <- ceiling(sd_calc)
print(sd_calc_ceil)

sd_calc_floor <- floor(sd_calc)
print(sd_calc_floor)

sd_calc_round <- round(sd_calc, digits = 4)
print(sd_calc_round)
```
</details>

Ok, now you're ready to learn about strings, let's go.